Decoding the Impact: Statistical Analysis of Factors Contributing to NFL Injuries

### 2024-04-24 paper.qmd
adding logistic regression tests

---
title: "Decoding the Impact: Statistical Analysis of Factors Contributing to NFL Injuries"
author: "Vincent Xie"
toc: false
number-sections: false
execute:
    echo: false
highlight-style: pygments
format: 
  html: 
    code-fold: true
    html-math-method: katex
  pdf: 
    fontsize: 12pt
    geometry: 
      - top=30mm
      - left=20mm
header-includes:
  - \usepackage{setspace}
  - \doublespacing
##  docx: default
---
```{python}
# Packages
import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns
import statsmodels.api as sm
from scipy.stats import chi2_contingency
```

@@

```{python}
# Load the datasets
injury = pd.read_csv('data/InjuryRecord.csv')
plays = pd.read_csv('data/PlayList.csv')
```

@@

# Bar chart for frequency of stadium types in PlayList
stadium_freq = plays['StadiumType'].value_counts()
plt.figure(figsize=(8, 4))
stadium_freq.plot(kind='bar', color='lightgreen')
plt.title('Frequency of Stadium Types')
plt.xlabel('Stadium Type')
plt.ylabel('Frequency')
plt.show()
```

@@

```{python}
# Bar chart for frequency of each injury type in InjuryRecord
injury_freq = injury['BodyPart'].value_counts()
plt.figure(figsize=(5, 2))
injury_freq.plot(kind='bar', color='skyblue')
plt.title('Frequency of Each Injury')
plt.xlabel('Body Part')
plt.ylabel('Frequency')
plt.show()

# Merge injury_record and playlist on PlayerKey
merged_data = pd.merge(injury, plays, on='PlayKey', how='inner')

# Calculate the frequency of injuries by roster position
injury_position_frequency = merged_data['RosterPosition'].value_counts()

# Plotting the frequency of injuries by roster position
plt.figure(figsize=(8, 4))  # Smaller plot size for better readability
injury_position_frequency.plot(kind='bar', color='salmon')
plt.title('Frequency of Injuries by Roster Position')
plt.xlabel('Roster Position')
plt.ylabel('Frequency')
plt.tight_layout()  # Adjust layout to fit elements
plt.show()
```

```{python}
# Merge injury_record and playlist on PlayerKey
merged_data = pd.merge(injury, plays, on='PlayKey', how='inner')

@@

# Plot a heatmap to visualize the frequency of each type of injury by each position group
plt.figure(figsize=(14, 8))
sns.heatmap(pivot_table, annot=True, cmap="viridis")
plt.title('Frequency of Each Type of Injury by Position Group')
plt.xlabel('Body Part')
plt.ylabel('Roster Position')
plt.show()
```

@@

```{python}
# Create a set of unique keys from InjuryRecord which indicates injuries
injury_keys = set(injury['PlayKey'])

@@

>>>>

---
title: "Decoding the Impact: Statistical Analysis of Factors Contributing to NFL Injuries"
author: "Vincent Xie"
toc: false
number-sections: false
execute:
    echo: false
highlight-style: pygments
format: 
  html: 
    code-fold: true
    html-math-method: katex
  pdf: 
    fontsize: 12pt
    geometry: 
      - top=30mm
      - left=20mm
header-includes:
  - \usepackage{setspace}
  - \doublespacing
##  docx: default
---
```{python}
# Packages
import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns
import statsmodels.api as sm
from scipy.stats import chi2_contingency
import numpy as np
import warnings

warnings.filterwarnings('ignore', category=FutureWarning)
```

@@

```{python}
# Load the datasets
injury = pd.read_csv('data/InjuryRecord.csv')
plays = pd.read_csv('data/PlayList.csv')
new_injury = pd.read_csv('data/mod_InjuryRecord.csv')
new_plays = pd.read_csv('data/mod_PlayList.csv')
```

@@

# Bar chart for frequency of stadium types in PlayList
stadium_freq = plays['StadiumType'].value_counts()
plt.figure(figsize=(6, 1))
stadium_freq.plot(kind='bar', color='lightgreen')
plt.title('Frequency of Stadium Types')
plt.xlabel('Stadium Type')
plt.ylabel('Frequency')
plt.show()
```

We can also group the temperatures and weather so make the data set more clean. Otherwise, there will be too many unique values. The labels for temperature are the following:

+ Cold: Below 50F
+ Cool: 50F-60F
+ Mild: 61F-70F
+ Warm: 71F-80F
+ Hot: Above 80F

For weather, it has been grouped to `rain`, `overcast`, `clear`, `snow`, and any others were grouped to `None` or `Unknown`.

```{python}
# Define the temperature bins and labels
bins = [-float('inf'), 50, 60, 70, 80, float('inf')]
labels = ['Cold', 'Cool', 'Mild', 'Warm', 'Hot']

# Replace the temperature values with categorical labels directly
plays['Temperature'] = pd.cut(plays['Temperature'], bins=bins, labels=labels)

# Save the modified dataframe if needed
plays.to_csv('data/mod_PlayList.csv', index=False)
```

```{python}
def map_weather(weather):
    if pd.isna(weather):
        return 'Unknown'
    weather = weather.lower()
    # Broadening the conditions for rain to capture "rainy", "showers", etc.
    if any(word in weather for word in ['rain', 'showers', 'rainy', 'rain shower']):
        return 'Rain'
    # Check for overcast and cloudy conditions
    elif any(word in weather for word in ['overcast', 'cloudy']):
        return 'Overcast'
    # Broadening conditions for clear to include "clear", "sunny", etc.
    elif any(word in weather for word in ['clear', 'sunny', 'sun']):
        return 'Clear'
    # Check for snow
    elif 'snow' in weather:
        return 'Snow'
    # Indoor conditions
    elif any(word in weather for word in ['indoor', 'indoors', 'controlled climate', 'n/a', 'none']):
        return 'None'
    return 'Unknown'

# Apply the mapping function to the Weather column
plays['Weather'] = plays['Weather'].apply(map_weather)

# Save the modified dataframe if needed
plays.to_csv('data/mod_PlayList.csv', index=False)
```

@@

From the collected the data, it is important to visualize the occurrences of injuries, especially the type and by the position. Below is a heat map that shows how many injury occurrences there are based on position. In terms of the most frequent injuries, there are many `Ankle` and `Knee` injuries, specifically in the `Linebacker` and `Wide Receiver` position groups. `Toes` tend to be the least frequent, with this data set only having one instance. This heatmap shows that with the playstyle of the sport, injured players tend to have knee and ankle issues.

```{python}
# Merge injury_record and playlist on PlayerKey
merged_data = pd.merge(new_injury, plays, on='PlayKey', how='inner')

@@

# Plot a heatmap to visualize the frequency of each type of injury by each position group
plt.figure(figsize=(5, 2))
sns.heatmap(pivot_table, annot=True, cmap="viridis")
plt.title('Frequency of Each Type of Injury by Position Group')
plt.xlabel('Body Part')
plt.ylabel('Roster Position')
plt.show()
```

```{python}
# Given the code snippets provided by the user, we will combine both plots into a single figure with subplots.

# First, let's recreate the datasets based on the previous analysis steps.
# NOTE: For the purposes of this demonstration, we will assume 'new_injury' and 'new_plays' are the injury_data and playlist_data respectively.

injury_counts = new_injury['Surface'].value_counts()
injury_percentages = (injury_counts / injury_counts.sum() * 100).round(2)

# Calculate the predominant surface for each game in the playlist data
predominant_surface_per_game = new_plays.groupby('GameID')['FieldType'].agg(pd.Series.mode).to_frame()
surface_type_counts = predominant_surface_per_game['FieldType'].apply(lambda x: x[0] if isinstance(x, list) else x).value_counts()
surface_type_percentages = (surface_type_counts / surface_type_counts.sum() * 100).round(2)

# Set up a figure with two subplots
fig, axs = plt.subplots(1, 2, figsize=(12, 6))  # Increased figure size for clarity

# Plot for injuries by surface type
axs[0].bar(injury_counts.index, injury_counts, color='skyblue')
axs[0].set_title('Number of Injuries by Surface Type')
axs[0].set_xlabel('Surface Type')
axs[0].set_ylabel('Number of Injuries')
axs[0].grid(axis='y', linestyle='--', alpha=0.7)

# Annotate percentages above the bars for the first plot
for i, v in enumerate(injury_counts):
    axs[0].text(i, v + 1, f'{injury_percentages.iloc[i]}%', ha='center', va='bottom')

# Plot for number of games played on each surface type
axs[1].bar(surface_type_counts.index, surface_type_counts, color=['green', 'blue'])
axs[1].set_title('Number of Games Predominantly Played on Each Surface Type')
axs[1].set_xlabel('Surface Type')
axs[1].set_ylabel('Number of Games')
axs[1].grid(axis='y', linestyle='--', alpha=0.7)

# Annotate percentages above the bars for the second plot
for i, v in enumerate(surface_type_counts):
    axs[1].text(i, v + 20, f'{surface_type_percentages.iloc[i]}%', ha='center', va='bottom')

# Set tight layout to prevent overlap
plt.tight_layout()

# Show the combined plot
plt.show()
```

```{python}
# To reflect the correct recovery time, we need to adjust the data.
# If there's a '1' in every 'DM' column for a record, the recovery is 42 days.
# If there are '1's in the first three 'DM' columns, recovery is 28 days, and so on.

# Create a new column that determines the maximum recovery time based on the 'DM' columns
def determine_recovery(row):
    if row['DM_M42'] == 1:
        return '42 Days'
    if row['DM_M28'] == 1:
        return '28 Days'
    if row['DM_M7'] == 1:
        return '7 Days'
    if row['DM_M1'] == 1:
        return '1 Day'
    return 'Unknown'

new_injury['RecoveryTime'] = new_injury.apply(determine_recovery, axis=1)

# Group by BodyPart and RecoveryTime to count the number of injuries
recovery_counts = new_injury.groupby(['BodyPart', 'RecoveryTime']).size().unstack(fill_value=0)

# Reorder the columns to match the desired recovery time order
recovery_counts = recovery_counts[['1 Day', '7 Days', '28 Days', '42 Days']]

# Calculate the percentage of recovery counts for each body part
recovery_percent = recovery_counts.div(recovery_counts.sum(axis=1), axis=0)

# Create the stacked bar chart with pastel colors
pastel_colors = ['#77dd77', '#fdfd96', '#84b6f4', '#ffbdbd']  # Pastel colors for green, yellow, blue, and pink respectively

# Plot the stacked bar chart with ordered recovery times
ax = recovery_percent.plot(kind='bar', stacked=True, figsize=(10, 6), color=pastel_colors)

# Annotate the counts on the bars with contrast adjustment for readability
for n, x in enumerate([*recovery_counts.index.values]):
    for (proportion, count, y_loc) in zip(recovery_percent.loc[x],
                                          recovery_counts.loc[x],
                                          recovery_percent.loc[x].cumsum()):
        # To ensure the text is readable, choose a text color with good contrast
        text_color = 'black'  # Black for contrast on pastel colors
        if count > 0:  # Only annotate non-zero counts
            ax.text(x=n, y=y_loc - proportion/2, s=int(count), ha='center', va='center', color=text_color)

ax.figure.gca().yaxis.set_major_formatter(plt.FuncFormatter(lambda x, _: f'{int(x*100)}%'))

plt.title('Proportion of Injury Types by Recovery Time')
plt.xlabel('Body Part Injured')
plt.ylabel('Proportion of Total Injuries')
plt.legend(title='Recovery Time', bbox_to_anchor=(1.04,1), loc="upper left")
plt.tight_layout()
plt.show()
```

@@

**Correlation between surface and injury occurrence**

```{python}
# Create a set of unique keys from InjuryRecord which indicates injuries
injury_keys = set(new_injury['PlayKey'])

@@

```{python}
# Merging the datasets on 'PlayerKey', 'GameID', and 'PlayKey'
merged_data = pd.merge(new_plays, new_injury, how='left', on=['PlayerKey', 'GameID', 'PlayKey'])

# Adding a binary column for whether an injury occurred
merged_data['InjuryOccurred'] = merged_data['BodyPart'].notna()

# Prepare the data for logistic regression
# We need to convert 'FieldType' into a binary variable for logistic regression
# Let's assume 'Synthetic' as 1 and 'Natural' as 0
merged_data['FieldTypeBinary'] = (merged_data['FieldType'] == 'Synthetic').astype(int)

# Dropping rows where 'FieldType' is NaN since we can't use them in the regression
merged_data_clean = merged_data.dropna(subset=['FieldType'])

# Logistic Regression: InjuryOccurred ~ FieldType
X = merged_data_clean[['FieldTypeBinary']]  # Independent variable
X = sm.add_constant(X)  # Adding a constant for the intercept term
y = merged_data_clean['InjuryOccurred']  # Dependent binary variable

# Fit the logistic regression model
logit_model = sm.Logit(y, X)
result = logit_model.fit()

# Display the summary of the logistic regression
result.summary()

```

**Weather conditions across different injury types**

```{python}
# Prepare the data for chi-square test on weather and injury types
# Drop rows where 'Weather' or 'BodyPart' is NaN since they can't be used in the test
weather_injury_data = merged_data.dropna(subset=['Weather', 'BodyPart'])

# Create the contingency table for 'Weather' and 'BodyPart'
contingency_weather_injury = pd.crosstab(weather_injury_data['Weather'], weather_injury_data['BodyPart'])

# Perform the chi-square test
chi2_weather_injury, p_value_weather_injury, dof_weather_injury, expected_weather_injury = chi2_contingency(contingency_weather_injury)

# Display the chi-square test results and the contingency table (if small enough)
contingency_weather_injury.shape, (chi2_weather_injury, p_value_weather_injury)

```

**Does temperature affect injury occurrence?**

```{python}
# Convert temperature categories into numerical codes for logistic regression
temperature_mapping = {'Cold': 0, 'Cool': 1, 'Mild': 2, 'Warm': 3, 'Hot': 4}
merged_data['TemperatureCode'] = merged_data['Temperature'].map(temperature_mapping)

# Prepare data for logistic regression
# Dropping rows where 'TemperatureCode' is NaN
temperature_injury_data = merged_data.dropna(subset=['TemperatureCode'])

# Logistic Regression: InjuryOccurred ~ Temperature
X_temp = temperature_injury_data[['TemperatureCode']]  # Independent variable
X_temp = sm.add_constant(X_temp)  # Adding a constant for the intercept term
y_temp = temperature_injury_data['InjuryOccurred']  # Dependent binary variable

# Fit the logistic regression model
logit_model_temp = sm.Logit(y_temp, X_temp)
result_temp = logit_model_temp.fit()

# Display the summary of the logistic regression
result_temp.summary()

```
